---
title: "Raster Builder"
author: "Nick McManus"
date: "2023-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(terra)     #GIS
library(geobr)     #administrative boundaries for Brazil
library(tidyverse) #always
library(here)      #reading/writing data easier
```

# OVERVIEW:
***

This code generates the rasters needed as inputs for the Shiny app. This include rasters of projected cropland abandonment, biodiversity, and carbon, both globally and for Brazil. The resolution and spatial extent of all global rasters must match each other (as well as the Brazilian layers) to generate proper zonal statistics and visual overlay. Global rasters are all at 50km resolution, while Brazilian rasters are at 1km in a projection better suited for the country.

This markdown is broken down into GLOBAL and BRAZIL sections, where the respective rasters are generated. These processed rasters are populated in the `data/processed` directory of this repository. More information regarding data sources can be found within this repo's README file.



# GLOBAL:
***

## CROPLAND ABANDONMENT

The code that generates projected cropland abandonment globally in 2050 is found in the `abandoned-cropland` repository. The outputs from this code will be read in and reclassified here. 

### Read in data

First read in projected 2050 LULC from Chen et al. for each SSP.
```{r}
# SSP1
ssp1_crop_2050 <- rast(here('data','raw','abandoned_crop','SSP1_abandoned_cropland_2050_global.tif'))
# SSP2
ssp2_crop_2050 <- rast(here('data','raw','abandoned_crop','SSP2_abandoned_cropland_2050_global.tif'))
# SSP3  
ssp3_crop_2050 <- rast(here('data','raw','abandoned_crop','SSP3_abandoned_cropland_2050_global.tif')) 
# SSP4
ssp4_crop_2050 <- rast(here('data','raw','abandoned_crop','SSP4_abandoned_cropland_2050_global.tif'))
# SSP5
ssp5_crop_2050 <- rast(here('data','raw','abandoned_crop','SSP5_abandoned_cropland_2050_global.tif'))
```



### Change resolution

Our global abandonment layers are currently at 1km resolution. For processing speed and ease of visualization, we will change the resolution to 50km. These new 50km cells will represent "proportion of abandonment". 
```{r}
# create function that finds the mean abandonment across 50km,
# then changes 0 values to NA for better viz
agg_crop_fxn <- function(ssp) {
  ssp_50k <- terra::aggregate(ssp, fact = 50, fun = 'mean') %>% 
    terra::project('epsg:4326')

  ssp_50k[ssp_50k==0] <- NA
  return(ssp_50k)
}

# use function on each SSP
ssp1_50k <- agg_crop_fxn(ssp1_crop_2050)
ssp2_50k <- agg_crop_fxn(ssp2_crop_2050)
ssp3_50k <- agg_crop_fxn(ssp3_crop_2050)
ssp4_50k <- agg_crop_fxn(ssp4_crop_2050)
ssp5_50k <- agg_crop_fxn(ssp5_crop_2050)
```


### Export Rasters
```{r}
writeRaster(ssp1_50k, 
            here('data', 'processed', 'global', 'ssp1_abandonment_global_50km.tif'), 
            overwrite = TRUE)
writeRaster(ssp2_50k, 
            here('data', 'processed', 'global', 'ssp2_abandonment_global_50km.tif'), 
            overwrite = TRUE)
writeRaster(ssp3_50k,
            here('data', 'processed', 'global', 'ssp3_abandonment_global_50km.tif'), 
            overwrite = TRUE)
writeRaster(ssp4_50k, 
            here('data', 'processed', 'global', 'ssp4_abandonment_global_50km.tif'),
            overwrite = TRUE)
writeRaster(ssp5_50k, 
            here('data', 'processed', 'global', 'ssp5_abandonment_global_50km.tif'), 
            overwrite = TRUE)
```



## BIODIVERSITY

[Info on biodiversity data]. 
```{r}
### read in raw biodiversity file
bio <- rast(here('data', 'raw', 'sparc_conservationPriorities.tif')) %>% 
  project(y = crs(ssp1_50k))

### change resolution and spatial extent to match abandonment
# to figure out factor that gets res closest to target raster, use this:
factor <- ceiling(res(ssp1_50k)[1]/res(bio)[1])

# get res as close as possible with factor
bio_agg <- aggregate(bio, fact = 4, fun = 'mean', na.rm = TRUE)

# use resample to make resolutions/spat extent EXACTLY the same
bio_50k <- resample(bio_agg, ssp1_50k, method = 'bilinear')

### export raster
writeRaster(bio_50k, 
            here('data','processed','global','biodiversity_global_50km.tif'),
            overwrite = TRUE)
```



## CARBON

This dataset, originally created by Cook-Patton et al. and updated by Global Forest Watch, estimates the rate of carbon sequestration in aboveground and belowground live biomass during the first 30 years of natural forest regeneration. Spatial sequestration estimates include all forest and savanna biomes in units of MgC/ha/yr at a ~1 km resolution. 

Source: https://data.globalforestwatch.org/documents/gfw::carbon-accumulation-potential-from-natural-forest-regrowth-in-forest-and-savanna-biomes/about

```{r}
### read in raw carbon file
carbon <- rast(here('data','raw','sequestration_rate_mean_aboveground_full_extent_Mg_C_ha_yr.tif')) %>% 
  project(y = crs(ssp1_50k))

# find factor for aggregation
factor <- ceiling(res(ssp1_50k)[1]/res(carbon)[1])

# lower resolution to 25km
carbon_agg <- aggregate(carbon, fact = 23, fun = 'mean', na.rm = TRUE)

# match spatial extent exactly
carbon_50k <- resample(carbon_agg, ssp1_50k, method = 'bilinear')

# export raster
writeRaster(carbon_50k, 
            here('data','processed','global','carbon_global_50km.tif'),
            overwrite = TRUE)
```







# BRAZIL:
***


## CROPLAND ABANDONMENT

The code that generates projected Brazilian cropland abandonment by 2050 is found in the `abandoned-cropland` repository. The outputs from this code will be read in and reclassified here. These are already in the correct resolution and spatial extent. They are used as reference for the biodiversity and carbon layers.

```{r}
# read in the cropland abandonment
ssp1_brazil <- rast(here('data','raw','abandoned_crop','SSP1_abandoned_cropland_brazil.tif'))
```


## BRAZIL BOUNDARIES

The boundary information for Brazilian biomes is read-in using the geobr package. Biome polygons are sourced from the Brazilian Institute of Geography and Statistics (IBGE). This dataset uses 2019 IBGE data at scale 1:250.000. Because carbon data for the Pantanal biome in Brazil is missing, this biome has been excluded from other rasters for restoration prioritization. 

Source: https://www.ibge.gov.br/geociencias/cartas-e-mapas/informacoes-ambientais/15842-biomas.html?=&t=acesso-ao-produto

```{r}
# read in biome boundaries from geobr package
biomes <- read_biomes(
  year = 2019,         #the most recent data available
  simplified = FALSE   #full resolution for boundaries
)

# create vector for whole of Brazil
biomes_vect <- biomes[-7, ] %>%  #remove coastal system biome
  st_transform(crs = crs(ssp1_brazil)) %>% #change crs to match other rasters
  vect()  #turn sf into vector; easier to crop rasts with

# create vector for Pantanal biome
pantanal <- biomes %>% 
  filter(code_biome == 6) %>% 
  st_transform(crs = crs(ssp1_brazil)) %>% 
  vect()

```


## BIODIVERSITY

After reading in the global layer, we will crop it to Brazil and resample to match the resolution of the planning units. Then, we will reclassify the layer to extinction risk values; this will help set tangible goals of minimizing extinction risk. 

```{r}
# read in global biodiversity data
bio_global <- rast(here("data","raw","sparc_conservationPriorities.tif"))

#temporarily change biome crs to crop/mask biodiversity data
#faster than changing crs of global raster
brazil_trans <- biomes_vect %>% 
  project(y = crs(bio_global))

# crop biodiversity layer to Brazil
bio_crop <- bio_global %>% 
  crop(brazil_trans) %>% 
  mask(brazil_trans) %>% 
  #change crs back to match other rasters
  terra::project(y = crs(biomes_vect))


# change resolution of biodiversity layer (4.72km) to match other rasters (1km)
# first use terra:disagg()
bio_disagg <- disagg(bio_crop, 
                     fact = 5,  
                     method = "bilinear")  

# resample higher res biodiversity to ensure same res and geometries 
# (fine-tuning the res)
bio_brazil <- resample(bio_disagg, ssp1_brazil, method = "bilinear")

# remove Pantanal biome from biodiversity layer
bio_brazil <- mask(bio_brazil, pantanal, inverse = TRUE, updatevalue = NA)

# change the name saved within raster for easier recall
# after stacking feature rasters
names(bio_brazil) <- "biodiversity_brazil"
```

Now reclassify Brazilian biodiversity according to extinction risk
```{r}
## read in values
extrisk <- read_csv(here("data", "raw", "globalverts_extrisk_scale.csv"))

## make three columns for a "from to" reclass
extrisk_df <- extrisk %>%
  # bring rasterval from previous row down as lower bound
  mutate(from = lag(rasterval), .before = 1) %>% 
  # rename rasterval to indicate upper bound
  rename(to = rasterval) %>% 
  # remove resulting na
  drop_na 

# turn csv into matrix
extrisk_m <- data.matrix(extrisk_df, rownames.force = NA)

# reclassify raster to extrisk matrix
bio_ext_brazil <- classify(bio_brazil, extrisk_m, right = TRUE)

## save new raster
writeRaster(bio_ext_brazil, 
            here("data","processed","brazil","biodiversity_extrisk_brazil_noPant.tif"),
            overwrite = TRUE)
```


## CARBON

Global carbon sequestration rates will be cropped and masked to Brazil, then resampled to ensure the same spatial extent as the cropland abandonment rasters. 
```{r}
# read in carbon layer
carbon_global <- rast(here("data", "raw", "sequestration_rate_mean_aboveground_full_extent_Mg_C_ha_yr.tif"))

#cannot crop/mask carbon data when biomes are different crs
#change biome crs because it's faster than changing crs of global raster
brazil_trans <- biomes_vect %>% 
  project(y = crs(carbon_global))

# crop carbon data to Brazil
carbon_brazil <- carbon_global %>% 
  crop(brazil_trans) %>% 
  mask(brazil_trans) %>%  
  #change crs back to match other rasters
  terra::project(y = crs(biomes_vect)) 

# default resolution of carbon layer is 898.8m
# change resolution to match other rasters (1km)
carbon_brazil <- resample(carbon_brazil, ssp1_brazil, method = "bilinear")

# remove Panatnal biome from carbon layer
carbon_brazil <- mask(carbon_brazil, pantanal, inverse = TRUE, updatevalue = NA, touches = TRUE)

# change the name saved within raster for easier recall
# after stacking feature rasters
names(carbon_brazil) <- "carbon_brazil"

# test plot to visualize results
plot(carbon_brazil)
plot(biomes_vect, add = TRUE)

# export 
writeRaster(carbon_brazil, 
            here("data","processed","brazil","carbon_brazil_noPant.tif"), 
            overwrite = TRUE)
```





